Discussion Draft

### 2020-04-16 doc/Discussion.rmd
adding in edits

We examine three modifier models and two functional and cell physiology models (table X) to test how well existing models fit out results. While there are challenges to fitting models built under different assumptions, we feel it's a useful exercise to highlight any unifying patterns across models.The three modifier models were built to explain variation in genome wide recombination rates between sexes (heterochiasmy), while the COM model was built to describe the interference pattern and difference in the recombination landscape between the sexes. In cases where the models can't be extended to predicting results, we list a 'NA' for the prediction.

@@

In our results and more generally in the literature, female biased heterochiasmy is the most common pattern observed, however male biased heterochiasmy is not especially rare [@brandvain2012scrambling]. Our results of male biased heterochiasmy in 2 strains is a novel finding due to the shorter evolutionary distances between the examined strains. The direction of heterochiasmy under the two locus modifier model is dependent on female specific recombination modifier suppresses or increases recombination, if the modifier is linked to the driving locus, and the meiotic stage (reduction or equatorial division) that the driver acts (see Table 1 of @brandvain2012scrambling).

Under the spindle based model, faster male evolution of the genome wide recombination rate, is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. <We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.> The SAC prevents anueploidy by ensuring all bivalents are correctly attached to the microtubulue, (bi-orientated) spindle before starting the irreversible metaphase-to-anaphase transition (cite).  A stable chromosome structure, through bi-orientation, requires, i) kinetochores attached to opposite poles of the cells and ii) at least one crossover to create tension across a proportion of sister cohesion (cite). Unstable bivalents on the meiosis I spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize on the microtubule spindle. If they fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents (such as those affecting recombination rate or cohesion), will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle. Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis ([@laneKauppi2019], Subrahman). 
These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volume (cite). This model nominates conserved differences in in the context of meiosis between males and females as the primary source of distinct evolutionary trajectories in genome wide recombination rates.

Our results extend previous reports [@dumont2011] of rapid male-specific evolution in multiple subspecies and strains of house mouse. Whether these instances of evolution of genome-wide recombination rates are due to independent events or segregating genetic variation within house mouse subspecies requires further study. Regardless, the faster male evolution in genome-wide recombination rates suggests that even for species with female biased heterochiasmy, selecting for higher male specific genome wide recombination rates may be the most effective way to increase the sex-averaged recombination rate for a genetic background. This insight could be important for animal breeding programs [@battagin2016].

@@

<While the modifier models parameterize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant.>

In the spindle based selection model, the less efficient SAC in oocytes leads to relaxed selection on the metaphase I chromosome structure. This will result in increased variance in chromosome structure and crossover number per cell, not just across strains, but across cells within individuals. Oocytes at metaphase I, have an acentriolar spindle comprised of  diffuse network of microtubules with multiple microtubule organizing centers [@schuh2007]. This is another way the spindle based selection model is linked to conserved sex differences in gametogenesis (cite).

@@

As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait. The two locus modifier and SACE models predict sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively.

The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differing strengths of chromosome movements during the pairing process [@hulten2011_COM] <@rubin2020mixing>.

Under the spindle based model we hypothesize that the sexual dimorphism in the recombination landscapes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. The irreversible process of the metaphase to anaphase transition is initiated by the protyltic decay of the sister cohesion connecting homologs (cite). The number and placement of crossovers alter the distribution of sister cohesion and the resulting chromosome structure when bivalents are aligned and bi-oriented on the metaphase plate (cite). Different timing of cell cycle between oogenesis and spermatogenesis may impose the the different selective pressures on sister cohesion. Faster spermatogenesis may select for synchronization of the separation homologs. While in oogenesis, the slower cell cycle and multiple arrest stages may require chromosome structures with greater stability on the MI spindle.

<The male crossover telomere bias could be adaptive in order to synchronize (or minimize) the timing of homologs separating at metaphase I>

@@

Our results from musculus^MSM^ and musculus^PWD^ show that the SC length and chromatin compaction is uncoupled from the direction of heterochiasmy. These results slightly depart from predictions which nominate chromatin organization (axis length) as the primary driver of heterochiasmy [@petkov2007]. While the relationship may hold for more general heterochiasmy patterns across broad scale comparisons, our results indicate the sexual dimorphism in chromatin compaction is conserved in house mouse (e.g. females have longer SC than males) and that the rapid male-specific evolution in crossover number per cell proceeded through another aspect of crossover regulation.

>>>>

To place our results in the broader context of meiotic variation we examine three modifier models and two functional and cell physiology models (table X). While there are challenges to fitting models built under different assumptions, we feel it's a useful exercise to highlight any unifying patterns for our specific traits across exsisting models. The three modifier models were built to explain variation in genome wide recombination rates between sexes (heterochiasmy), while the chromosome ocillatory movement (COM) model was built to describe positive interference. This spindle based selection model was developed to explain the results resented in this manuscript. In cases where the models can't be extended to predicting results, we list a 'NA' for the prediction.

@@

The direction of heterochiasmy under the two locus modifier model is dependent on female specific recombination modifier suppresses or increases recombination, if the modifier is linked to the driving locus, and the meiotic stage (reduction or equatorial division) when the driver acts (see Table 1 of @brandvain2012scrambling).

Under the spindle based model, faster male evolution of the genome wide recombination rate, is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. <We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.> The SAC prevents anueploidy by ensuring all bivalents are correctly attached to the microtubulue, (bi-orientated) spindle before starting the metaphase-to-anaphase transition ([@laneKauppi2019], subramanian and Hochwagen , @dumontDesai2012).

A stable chromosome structure requires, i) kinetochores attached to opposite poles of the cells and ii) at least one crossover to create tension across a proportion of sister cohesion [@laneKauppi2019]. Unstable bivalents on the meiosis I spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize on the microtubule spindle. If they fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents (such as those affecting recombination rate or cohesion), will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle. Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis ([@laneKauppi2019], Subrahman). These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle (@So2019, @dumontDesai2012) and cell volume [@kyogoku2017]. This model nominates conserved differences in in the context of meiosis between males and females as the primary source of distinct evolutionary trajectories in genome wide recombination rates.

In our results and more generally in the literature, female biased heterochiasmy is the most common pattern, however male biased heterochiasmy is not especially rare [@brandvain2012scrambling]. Our results extend previous reports of male biased heterochiasmy [@dumont2011] to  new subspecies of house mouse (*molossinus*) and emphasize the short evolutionary period. Whether these instances of evolution of genome-wide recombination rates are due to independent events or segregating genetic variation within house mouse subspecies requires further study. Regardless, the faster male evolution in genome-wide recombination rates suggests that even for species in female biased heterochiasmy species, selecting for higher male specific genome wide recombination rates may be the most effective way to increase the sex-averaged recombination rate for a genetic background. This insight could be important for animal breeding programs [@battagin2016].

@@

In the spindle based selection model, the less efficient SAC in oocytes leads to relaxed selection on the metaphase I chromosome structure. This will result in increased variance in chromosome structure and crossover number per cell, not just across strains, but across cells within individuals. Oocytes at metaphase I, have an acentriolar spindle comprised of  diffuse network of microtubules with multiple microtubule organizing centers [@schuh2007]. Centrosomes spindles are in spermatogensis and mature sperm cells delivers centrioles (most mammals) or centrosome material (rodents and rabbits) to the egg during fertilization (@manandhar2005centrosome). This is another way the spindle based selection model is linked to conserved sex differences in gametogenesis [@ross2015].

@@

As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait. The two locus modifier and SACE models predict sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively. The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differing strengths of chromosome movements during the pairing process [@hulten2011_COM].

Under the spindle based model we hypothesize that the sexual dimorphism in the recombination landscapes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. The irreversible process of the metaphase to anaphase transition is initiated by the protyltic decay of the sister cohesion connecting homologs [@laneKauppi2019], subramanian and Hochwagen , @dumontDesai2012]. The number and placement of crossovers alter the distribution of sister cohesion and the resulting chromosome structure when bivalents are aligned and bi-oriented on the metaphase plate [@vanVeen2003, @altendorfer2020]. Differences in timing of cell cycle between oogenesis and spermatogenesis imposes the the different selective pressures on how sister cohesion affects chromosome structure. Faster spermatogenesis may select for synchronization of the separation homologs (cite?). While in oogenesis, the slower cell cycle and multiple arrest stages may require chromosome structures with greater stability on the MI spindle, espicially for those with dicyate arrest [@lee2019].

@@

Our results from musculus^MSM^ and musculus^PWD^ demonstrate that the SC length/chromatin compaction are uncoupled from the direction of heterochiasmy in house mice. These results slightly depart from predictions which nominate chromatin compaction/organization (axis length) as the primary driver of heterochiasmy [@petkov2007].

The relationship may hold for more general heterochiasmy patterns across sexes but fail to fully explain rapid evolution in the sex specific context.  (this can be attributed to distinct evolutionary trajectories due to sex specific meiotic context)

Our results indicate the sexual dimorphism in chromatin compaction is conserved in house mouse (e.g. females have longer SC than males) and that the rapid male-specific evolution in crossover number per cell proceeded through another aspect of crossover regulation.
